|

### 2020-01-03 appendix/appendix.Rmd
version 0.2.1, small edits

---
title: |
       | Supplementary Materials for "Bottom-Up Accountability and Public Service Provision: Evidence from a Field Experiment in Brazil"
author:
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
date: \today
abstract: "Does local oversight improve public service delivery? We study the effect of a mobile phone application that allows citizens to monitor school construction projects in Brazilian municipalities. The app prompts users to submit data about construction sites, sends such crowdsourced information to independent engineers, and contacts the mayors' offices about project delays. Our results show that the app has a null impact on the school construction indicators. Additionally, we find that politicians are unresponsive to individual requests. The results question the impact of bottom-up monitoring on public service performance and suggest that interventions targeted to other groups, or focusing on different issues, may produce better policy outcomes."
abstractspacing: double
keywords: accountability, Brazil, impact evaluation, state capacity, technology
fontfamily: libertine
fontawesome: yes
fontsize: 12pt
monospace-url: yes
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    toc: true
    keep_tex: no
    template: template.latex
---

@@

The treatment is a compounded treatment comprised of steps (2) to (4). To
make sure that the only source of effect comes from step (2), we took the
following precautions:

@@

3.2. After we had the EWB report, we notified the mayor's office in compliance
with Brazil's Freedom of Information Act (FOI). We communicated that a citizen
had anonymously made a denounce about a school (fully identifying the school
construction that we were referring to), and that the Engineers without Borders
found the construction delayed. We submitted a request for the mayor's office
to provide an explanation on why the construction was delayed, and asked for
updated statistics on when the construction will be finished. According to the
FOI law, the mayor had twenty days to respond.

@@

We argue that the app promotes two crucial features of bottom-up
accountability. First, it helps people to access information that would
otherwise be available only in an obscure spreadsheet on the Ministry of
Education website. Second, whenever people find this information troublesome,
we are providing an easy way to them to act upon the information. As the _Tá de
Pé_ design was intuitive and straightforward, we believe that most of
Brazilians would find it easy to use the app.

Moreover, the success of treatment does not require that everyone in the
municipality uses it. The more usage means more pictures and denounces, and
this represents more pressure over the mayors office, what could revert into a
higher effect. However, just one denounce would be enough for the project to be
successful, as the denounce can be directed to the Brazilian General
Comptroller (CGU), and this could potentially deny future funding to the
municipality.

@@

1. Percentage of the project completed before the impact evaluation started:
	 This is a column in the Ministry of Education dataset. The data in this
	 column were sampled in the beginning of the intervention.

@@

All these variables come from our manipulation of the Ministry of Education
data. Collecting the outcomes in this way is superior than the alternative of
gathering the data ourselves. First, we would not be able to manipulate the
results and cherry-pick favourable results. Even if they have the best
intentions, researchers may bias the results just by the choices they make in
the data collection process. For instance, if we send an RA to collect the
data, as she knows the app, she could unconsciously over-report positive
results. This would bias our findings and undermine the credibility of our
experiment. Second, using external data makes the results easily reproducible
by other researchers. As the Ministry of Education has standardised data
collection procedures, they greatly simplify and facilitate the reproduction of
our findings.

@@

The app by itself would probably not have had an effect on the responsiveness
of mayors. To boost usage and publicize the app, Transparência Brasil ran 64
Facebook campaigns throughout the country. The Facebook campaign concentrates
the bulk of the investments in the beginning of the intervention. The app
download data is in the second panel of Figures 2 and 3. For the Facebook
interaction, we have the following statistics:

1. Impressions
- Mean: `r format(round(mean(fballt$impressoes),2), big.mark = ",")`
	- Total for all campaigns: `r format(sum(fballt$impressoes), big.mark = ",")`

>>>>

---
title: |
       | Supplementary Materials for "Bottom-Up Accountability and Public Service Provision: Evidence from a Field Experiment in Brazil"
author:
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
date: \today
abstract: "Does local oversight improve public service delivery? We study the effect of a mobile phone application that allows citizens to monitor school construction projects in Brazilian municipalities. The app prompts users to submit data about construction sites, sends such crowdsourced information to independent engineers, and contacts the mayors' offices about project delays. Our results show that the app has a null impact on school construction indicators. Additionally, we find that politicians are unresponsive to individual requests. The results question the impact of bottom-up monitoring on public service performance and suggest that interventions targeted at other groups, or focused on different issues, may produce better policy outcomes."
abstractspacing: double
keywords: accountability, Brazil, impact evaluation, state capacity, technology
fontfamily: libertine
fontawesome: yes
fontsize: 12pt
monospace-url: yes
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    toc: true
    keep_tex: no
    template: template.latex
---

@@

The treatment is a compounded treatment comprised of steps (2) to (4). To
make sure that the only source of the effect comes from step (2), we took the
following precautions:

@@

3.2. After we had the EWB report, we notified the mayor's office in compliance
with Brazil's Freedom of Information Act (FOI). We communicated that a citizen
had anonymously made a denounce about a school (fully identifying the school
construction that we were referring to), and that the Engineers without Borders
found the construction delayed. We submitted a request for the mayor's office
to provide an explanation on why the construction was delayed and asked for
updated statistics on when the construction will be finished. According to the
FOI law, the mayor had twenty days to respond.

@@

We argue that the app promotes two crucial features of bottom-up
accountability. First, it helps people to access information that would
otherwise be available only in an obscure spreadsheet on the Ministry of
Education website. Second, whenever people find this information troublesome,
we are providing an easy way for them to act upon the information. As the _Tá
de Pé_ design was intuitive and straightforward, we believe that most
Brazilians would find it easy to use the app.

Moreover, the success of treatment does not require that everyone in the
municipality uses it. The more usage means more pictures and denounces, and
this represents more pressure over the mayors' office, which could revert into
a higher effect. However, just one denounce would be enough for the project to
be successful, as the denounce can be directed to the Brazilian General
Comptroller (CGU), and this could potentially deny future funding to the
municipality.

@@

1. Percentage of the project completed before the impact evaluation started:
	 This is a column in the Ministry of Education dataset. The data in this
	 column were sampled at the beginning of the intervention.

@@

All these variables come from our manipulation of the Ministry of Education
data. Collecting the outcomes in this way is superior to the alternative of
gathering the data ourselves. First, we would not be able to manipulate the
results and cherry-pick favourable results. Even if they have the best
intentions, researchers may bias the results just by the choices they make in
the data collection process. For instance, if we send an RA to collect the
data, as she knows the app, she could unconsciously over-report positive
results. This would bias our findings and undermine the credibility of our
experiment. Second, using external data makes the results easily reproducible
by other researchers. As the Ministry of Education has standardised data
collection procedures, they greatly simplify and facilitate the reproduction of
our findings.

@@

The app by itself would probably not have had an effect on the responsiveness
of mayors. To boost usage and publicize the app, Transparência Brasil ran 64
Facebook campaigns throughout the country. The Facebook campaign concentrates
the bulk of the investments at the beginning of the intervention. The app
download data is in the second panel of Figures 2 and 3. For the Facebook
interaction, we have the following statistics:

1. Impressions
	- Total for all campaigns: `r format(sum(fballt$impressoes), big.mark = ",")`

### 2020-01-03 article/article.Rmd
version 0.2.1, small edits

---
title:  |
        | Bottom-Up Accountability and Public Service Provision: Evidence from a Field Experiment in Brazil
author:
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
date: \today
abstract: "Does local oversight improve public service delivery? We study the effect of a mobile phone application that allows citizens to monitor school construction projects in Brazilian municipalities. The app prompts users to submit data about construction sites, sends such crowdsourced information to independent engineers, and contacts the mayors' offices about project delays. Our results show that the app has a null impact on the school construction indicators. Additionally, we find that politicians are unresponsive to individual requests. The results question the impact of bottom-up monitoring on public service performance and suggest that interventions targeted to other groups, or focusing on different issues, may produce better policy outcomes."
abstractspacing: double
keywords: accountability, Brazil, impact evaluation, state capacity, technology
fontsize: 12pt
margin:
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    template: article-template.latex
---

@@

We use the TDP app to conduct two experimental interventions and test the
impact of citizen oversight on five outcomes related to school completion rates
and complaints to public authorities. Overall, providing information to
citizens has no consistent impact on policy outcomes. In the first experiment,
we find that the TDP app increased the likelihood of construction cancellation
by 2 percent, but the result does not replicate. The remaining five models have
null results. In our second intervention, none of the estimations reach
conventional levels of statistical significance. Importantly, all coefficients
are small, what suggests that even if the TDP app had a significant effect on
the outcomes, its substantive impact would be negligible.

@@

We also note that effect sizes are small in all estimations and that the
coefficients flip signs in all but one of our six dependent variables. This
provides further evidence for the null results: not only the app would have
a low impact even if the treatment were significant, but the results could go
against what the bottom-up accountability theory predicts. At least in the
school construction outcomes we investigate here, we find little evidence that
local monitoring works in the context of school constructions in Brazil.

@@

In this paper, we discuss a mobile phone application fosters community
oversight and political accountability in Brazil. Our two interventions show
that the results are at best mixed. Although we find some treatment effect on
school cancelling rates in the first intervention, the app has no consistent
impact on our outcomes of interest. These findings add to the studies that cast
doubts on the relationship between bottom-up accountability and local policy
performance [e.g., @banerjee2010pitfalls; @lieberman2014does;
@raffler2018weakness].

What factors, then, are driving these results? It seems unlikely that the null
results derive from flaws in the research design. First, our study is well
powered. Although the treatment is indirect---the person has to download the
app, find a school construction, and then report it---,we included
a substantial number of schools in the treatment groups. Second, balance and
manipulation tests indicate that the treatment allocation was successful, so we
can rule out problems in the randomisation procedures. Third, data from
Google Analytics confirm that citizens indeed used the app and provided
information to our dataset. This indicates that the treatment manipulation was
effective. In this sense, it is unlikely that our results derive from low user
response. Fourth, after doing a series of robustness tests, we still find no
firm evidence of treatment effect. Finally, note that the sign of the
coefficients are frequently contrary to our theoretical expectations. This
rules out a possible concern about statistical power with our small control
group approach.

We discuss some possible reasons why community monitoring did not work in our
case. One plausible explanation is that individuals were unable to
differentiate the effect of political corruption from those of spending cuts.
Due to the severe economic crisis in 2014--2016, the Brazilian federal
government introduced discretionary spending limits that affected public
investment [@rossi2016impactos]. Politicians may argue that delays in school
constructions are not derived from their misuse of government funds, but from
the austerity measures. If this is the case, citizens will not blame local
politicians for the underprovision of public goods. Consequently,
representatives can dismiss individual requests as the issue is unlikely to
escalate.

The electoral cycle might also have decreased the potential effect of the
treatment. As the experiment was fielded right after Brazil's municipal
elections, incumbents might have disregarded the requests because they did not
see the demands as politically costly in the short run. Having just taken
office, mayors might have focused their attention to the formation of
government coalitions or to budget concerns. Future research may evaluate how
electoral dynamics interact with citizen oversight, potentially by replicating
informational experiments in different stages of the political cycle.

@@

\newpage
\setlength{\parindent}{0cm} 
\setlength{\parskip}{5pt}

>>>>

---
title:  |
        | Bottom-Up Accountability and Public Service Provision: Evidence from a Field Experiment in Brazil
author:
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
date: \today
abstract: "Does  local oversight improve public service delivery? We study the effect of a mobile phone application that allows citizens to monitor school construction projects in Brazilian municipalities.  The app prompts users to submit data about construction sites, sends such crowdsourced information to independent engineers, and contacts the mayors' offices about project delays. Our results show that the app has a null impact on school construction indicators. Additionally, we find that politicians are unresponsive to individual requests. The results question the impact of  bottom-up monitoring on public service performance and suggest that interventions targeted at other groups, or focused on different issues, may produce better policy outcomes."
abstractspacing: double
keywords: accountability, Brazil, impact evaluation, state capacity, technology
fontsize: 12pt
margin:
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    template: article-template.latex
---

@@

We use the TDP app to conduct two experimental interventions and test the
impact of citizen oversight on five outcomes related to school completion rates
and complaints to public authorities. Overall, providing information to
citizens has no consistent impact on policy outcomes. In the first experiment,
we find that the TDP app increased the likelihood of construction cancellation
by 2 per cent, but the result does not replicate. The remaining five models
have null results. In our second intervention, none of the estimations reach
conventional levels of statistical significance. Importantly, all coefficients
are small, what suggests that even if the TDP app had a significant effect on
the outcomes, its substantive impact would be negligible.

@@

We also note that effect sizes are small in all estimations and that the
coefficients flip signs in all but one of our six dependent variables. This
provides further evidence for the null results: not only the app would have
a low impact even if the treatment were significant, but the results could go
against what the bottom-up accountability theory predicts. At least in the
school construction outcomes we investigate here, we find little evidence that
grassroots monitoring works in the context of school constructions in Brazil.

@@

In this paper, we discuss whether delivering information to citizens via a
mobile phone application fosters community oversight and political
accountability in Brazil. Our two interventions show that the results are at
best mixed. Although we find some treatment effect on school cancelling rates
in the first intervention, the app has no consistent impact on our outcomes of
interest. These findings add to the studies that cast doubts on the
relationship between bottom-up accountability and local policy performance
[e.g., @banerjee2010pitfalls; @lieberman2014does; @raffler2018weakness].

What factors, then, are driving these results? It seems unlikely that the null
results derive from flaws in the research design. First, our study is well
powered. Although the treatment is indirect---the person has to download the
app, find a school construction, and then report it---, we included a
substantial number of schools in the treatment groups. Second, balance and
manipulation tests indicate that the treatment allocation was successful, so we
can rule out problems in the randomisation procedures. Third, data from
Google Analytics confirm that citizens indeed used the app and provided
information to our dataset. This indicates that the treatment manipulation was
effective. In this sense, it is unlikely that our results derive from low user
response. Fourth, after doing a series of robustness tests, we still find no
firm evidence of treatment effect. Finally, note that the signs of the
coefficients are frequently contrary to our theoretical expectations. This
rules out a possible concern about statistical power with our small control
group approach.

We discuss some possible reasons why community monitoring did not work in our
case. One plausible explanation is that individuals were unable to
differentiate the effect of political corruption from those of spending cuts.
Due to the severe economic crisis in 2014--2016, the Brazilian federal
government introduced discretionary spending limits that affected public
investment [@rossi2016impactos]. Politicians may argue that delays in school
constructions are not derived from their misuse of government funds but from
the austerity measures. If this is the case, citizens will not blame local
politicians for the underprovision of public goods. Consequently,
representatives can dismiss individual requests as the issue is unlikely to
escalate.

The electoral cycle might also have decreased the potential effect of the
treatment. As the experiment was fielded right after Brazil's municipal
elections, incumbents might have disregarded the requests because they did not
see the demands as politically costly in the short run. Having just taken
office, mayors might have focused their attention on the formation of
government coalitions or to budget concerns. Future research may evaluate how
electoral dynamics interact with citizen oversight, potentially by replicating
informational experiments in different stages of the political cycle.

@@

\setlength{\parindent}{0cm}
\setlength{\parskip}{5pt}
